---
title: "External validation of the performance of competing risks prediction models: a guide through modern methods -  Develop a risk prediction model using the subdistribution hazard approach"
always_allow_html: true
output:
  github_document:
    toc: true
    toc_depth: 4
  keep_text: true
  pandoc_args: --webtex
---
  

### Installing and loading packages and import data

The following libraries are needed to achieve the outlined goals, the code chunk below will a) check whether you already have them installed, b) install them for you if not already present, and c) load the packages into the session.


```{r setup, include=FALSE}
# Knitr options
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.path = "imgs/Development_SDH/"
)
```

```{r, wdlib, message=FALSE,warning=FALSE}
# Use pacman to check whether packages are installed, if not load
if (!require("pacman")) install.packages("pacman")
library(pacman)

pacman::p_load(
  survival,
  rms,
  mstate,
  pseudo,
  pec,
  riskRegression,
  plotrix,
  knitr,
  splines,
  kableExtra,
  gtsummary,
  boot,
  tidyverse,
  rsample,
  gridExtra,
  webshot
)

# Install latest development version of riskRegression
if (!require("devtools", character.only = TRUE)) install.packages("devtools")
if (!require("riskRegression", character.only = TRUE)) devtools::install_github("tagteam/riskRegression")
require("riskRegression", character.only = TRUE)

# Import data ------------------
rdata <- readRDS(here::here("Data/rdata.rds"))
vdata <- readRDS(here::here("Data/vdata.rds"))

rdata$hr_status <- relevel(rdata$hr_status, ref = "ER and/or PR +")
vdata$hr_status <- relevel(vdata$hr_status, ref = "ER and/or PR +")
```

We loaded the development data (rdata) and the validation data (vdata).
More details about development and validation data are provided in the manuscript.


### Descriptive statistics

```{r, import,echo=FALSE}
rsel <- rdata[, c("id", "age", "size", "ncat", "hr_status")]
vsel <- vdata[, c("id", "age", "size", "ncat", "hr_status")]
rsel$dt <- 1
vsel$dt <- 2
cdata <- rbind(rsel, vsel)
cdata$dt <- factor(cdata$dt,
                   levels = c(1, 2),
                   labels = c("Development data", "Validation data")
)
label(cdata$age) <- "Age"
label(cdata$size) <- "Size"
label(cdata$ncat) <- "Nodal status"
label(cdata$hr_status) <- "Hormon receptor status"
# Units
units(cdata$age) <- "years"
units(cdata$size) <- "cm"
```

```{r tab1_bis, echo=FALSE}
gtsummary::tbl_summary(
  data = cdata %>% select(-id),
  label = list(age ~ "Age (years)", size ~ "Size (cm)"),
  by = "dt", 
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({min}, {max})"
  ),
) %>% 
  gtsummary::as_kable_extra() %>% 
  kableExtra::kable_styling("striped") 
```

## Develop a competing risks prediction model using the subdistribution hazard approach


### Cumulative incidence curves
First, we draw the cumulative incidence curves of breast cancer recurrence.

```{r, cuminc, fig.align='center'}
# Expand datasets -------------------------
# Create indicator variables for the outcomes in the development set
rdata$status1[rdata$status_num == 1] <- 1
rdata$status1[rdata$status_num != 1] <- 0
rdata$status2[rdata$status_num == 2] <- 2
rdata$status2[rdata$status_num != 2] <- 0

# Create indicator variables for the outcomes in the validation set
vdata$status1[vdata$status_num == 1] <- 1
vdata$status1[vdata$status_num != 1] <- 0
vdata$status2[vdata$status_num == 2] <- 2
vdata$status2[vdata$status_num != 2] <- 0


# Expand data to prepare for fitting the model 
rdata.w <- crprep(
  Tstop = "time",
  status = "status_num",
  trans = c(1, 2),
  id = "id",
  keep = c("age", "size", "ncat", "hr_status"),
  data = rdata
)
# Save extended data with weights for recurrence (failcode=1)
# and non recurrence mortality (failcode=2)
rdata.w1 <- rdata.w %>% filter(failcode == 1)
rdata.w2 <- rdata.w %>% filter(failcode == 2)

vdata.w <- crprep(
  Tstop = "time",
  status = "status_num",
  trans = c(1, 2),
  id = "id",
  keep = c("age", "size", "ncat", "hr_status"),
  data = vdata
)
vdata.w1 <- vdata.w %>% filter(failcode == 1)
vdata.w2 <- vdata.w %>% filter(failcode == 2)

# Development set --------
mfit_rdata <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = rdata.w1, weights = weight.cens
)
mfit_vdata <- survfit(
  Surv(Tstart, Tstop, status == 1) ~ 1,
  data = vdata.w1, weights = weight.cens
)
par(xaxs = "i", yaxs = "i", las = 1)
oldpar <- par(mfrow = c(1, 2), mar = c(5, 5, 1, 1))
plot(mfit_rdata,
     col = 1, lwd = 2,
     xlab = "Years since BC diagnosis",
     ylab = "Cumulative incidence", bty = "n",
     ylim = c(0, 0.25), xlim = c(0, 5), fun = "event", conf.int = TRUE
)
title("Development data")
plot(mfit_vdata,
     col = 1, lwd = 2,
     xlab = "Years since BC diagnosis",
     ylab = "Cumulative incidence", bty = "n",
     ylim = c(0, 0.25), xlim = c(0, 5), fun = "event", conf.int = TRUE
)
title("Validation data")
par(oldpar)
# Cumulative incidences
smfit_rdata <- summary(mfit_rdata, times = c(1, 2, 3, 4, 5))
smfit_vdata <- summary(mfit_vdata, times = c(1, 2, 3, 4, 5))
```

The R packages and functions `cmprsk::cuminc()` and `mstate::Cuminc()`are good and easy alternatives to estimate the cumulative incidence function. 

```{r, res_ci, fig.align='center',echo=FALSE}
res_ci <- cbind(
  1 - smfit_rdata$surv,
  1 - smfit_rdata$upper,
  1 - smfit_rdata$lower,
  1 - smfit_vdata$surv,
  1 - smfit_vdata$upper,
  1 - smfit_vdata$lower
)
res_ci <- round(res_ci, 2)
rownames(res_ci) <- c(
  "1-year", "2-year",
  "3-year", "4-year",
  "5-year"
)
colnames(res_ci) <- rep(c(
  "Estimate", "Lower .95",
  "Upper .95"
), 2)
kable(res_ci,
      row.names = TRUE
) %>%
  kable_styling("striped", position = "center") %>%
  add_header_above(c(" " = 1, "Development data" = 3, "Validation data" = 3))
```

The 5-year cumulative incidence of breast cancer recurrence was 14% (95% CI: 11-16%), and 10% (95%CI: 8-12%)

### Check non-linearity of continuous predictors

Here we investigate the potential non-linear relation between continuous predictors (i.e. age and size) and the outcomes. We apply three-knot restricted cubic splines using `rms::rcs()` function (details are given in e.g. Frank Harrell's book 'Regression Model Strategies (second edition)', page 27.

```{r,ff, warning=FALSE, fig.align='center'}

# Defining knots of the restricted cubic splines ------------------
# Extract knots position of the restricted cubic spline based on the
# original distribution of the data

# Age at breast cancer diagnosis
rcs3_age <- rcspline.eval(rdata$age, nk = 3)
attr(rcs3_age, "dim") <- NULL
attr(rcs3_age, "knots") <- NULL
pos_knots_age <- attributes(rcspline.eval(rdata$age, nk = 3))$knots
rdata$age3 <- rcs3_age

# Size of breast cancer
rcs3_size <- rcspline.eval(rdata$size, nk = 3)
attr(rcs3_size, "dim") <- NULL
attr(rcs3_size, "knots") <- NULL
pos_knots_size <- attributes(rcspline.eval(rdata$size, nk = 3))$knots
rdata$size3 <- rcs3_size

# FG model
dd <- datadist(rdata)
options(datadist = "dd")
fit_fg_rcs <- cph(Surv(Tstart, Tstop, status == 1) ~ 
                    rcs(age, pos_knots_age) + rcs(size, pos_knots_size) +
                    ncat + hr_status, 
                  weights = weight.cens,
                  x = T, 
                  y = T, 
                  surv = T, 
                  data = rdata.w1
)
P_fg_age_rcs <- Predict(fit_fg_rcs, "age")
P_fg_size_rcs <- Predict(fit_fg_rcs, "size")
# print(fit_fg_rcs)
# print(summary(fit_fg_rcs))
# print(anova(fit_fg_rcs))

oldpar <- par(mfrow = c(1, 2), mar = c(5, 5, 1, 1))
par(xaxs = "i", yaxs = "i", las = 1)

# FG - age
plot(P_fg_age_rcs$age, 
     P_fg_age_rcs$yhat,
     type = "l", 
     lwd = 2, 
     col = "blue", 
     bty = "n",
  xlab = "Age at breast cancer diagnosis",
  ylab = "log Relative Hazard", 
  ylim = c(-2, 2), 
  xlim = c(65, 95)
)
polygon(c(P_fg_age_rcs$age, 
          rev(P_fg_age_rcs$age)),
  c(P_fg_age_rcs$lower, 
    rev(P_fg_age_rcs$upper)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(P_fg_age_rcs$age, 
     P_fg_age_rcs$yhat,
     type = "l", 
     lwd = 2, 
     col = "blue", 
     bty = "n",
     xlab = "Age at breast cancer diagnosis", 
     ylab = "log Relative Hazard",
     ylim = c(-2, 2),
     xlim = c(65, 95)
)

# FG - size
par(xaxs = "i", yaxs = "i", las = 1)
plot(P_fg_size_rcs$size, 
     P_fg_size_rcs$yhat,
     type = "l", 
     lwd = 2, 
     col = "blue", 
     bty = "n",
     xlab = "Size of breast cancer",
     ylab = "log Relative Hazard", 
     ylim = c(-2, 2),
     xlim = c(0, 7)
)
polygon(c(P_fg_size_rcs$size, 
          rev(P_fg_size_rcs$size)),
  c(P_fg_size_rcs$lower, 
    rev(P_fg_size_rcs$upper)),
  col = "grey75",
  border = FALSE
)
par(new = TRUE)
plot(P_fg_size_rcs$size, 
     P_fg_size_rcs$yhat,
     type = "l", 
     lwd = 2, 
     col = "blue", 
     bty = "n",
     xlab = "Size of breast cancer", 
     ylab = "log Relative Hazard",
     ylim = c(-2, 2), 
     xlim = c(0, 7)
)
par(oldpar)
options(datadist = NULL)

# Fit the Fine and Gray model assuming linear relation of
# the continuous predictors
dd <- datadist(rdata, adjto.cat = "first")
options(datadist = "dd")
fit_fg <- cph(Surv(Tstart, Tstop, status == 1) ~ age + size +
                ncat + hr_status,
              weights = weight.cens,
              x = T, 
              y = T, 
              surv = T, 
              data = rdata.w1
)
options(datadist = NULL)
```

```{r, res_aic, fig.align='center',echo=FALSE}
res_AIC <- matrix(c(AIC(fit_fg), 
                    AIC(fit_fg_rcs)),
                  byrow = T,
                  ncol = 2, 
                  nrow = 1,
  dimnames = list(
    c("Fine and Gray models"),
    c(
      "AIC without splines",
      "AIC with splines"
    )
  )
)
kable(res_AIC,
  row.names = TRUE
) %>%
  kable_styling("striped", position = "center")
```

The AIC and the graphical check suggested a potential linear relation between the continuous predictors (age and size) and the event of interest (breast cancer recurrence).

### Checking proportional subdistribution hazards assumption

We now examine the fits further by checking the proportionality of the subdistribution hazards of the models.


```{r,sph, message=FALSE, warning=FALSE,fig.align='center'}
zp_fg <- cox.zph(fit_fg, transform = "identity")

par(las = 1, xaxs = "i", yaxs = "i")
# c(bottom, left, top, right)
oldpar <- par(mfrow = c(2, 2), mar = c(5, 6.1, 3.1, 1))
sub_title <- c("Age", "Size", "Lymph node status", "HR status")
for (i in 1:4) {
  plot(zp_fg[i], 
       resid = F, 
       bty = "n", 
       xlim = c(0, 5))
  abline(0, 0, lty = 3)
  title(sub_title[i])
}
mtext("Fine and Gray", 
      side = 3, 
      line = -1, 
      outer = TRUE, 
      font = 2)
par(oldpar)

kable(round(zp_fg$table, 3)) %>%
  kable_styling("striped", position = "center")
```

The statistical tests showed a potential violation of the proportional hazards assumption for size of breast cancer in the model for non-recurrence mortality. For simplicity we ignore this violation in the remainder.


### Model development - fit the risk prediction models
We develop and show the results of the Fine and Gray subdistribution hazard regression model.
In R, there are multiple alternatives to fit a competing risks model using the subdistribution hazard approach:

+ Using `survival::coxph()` after using `survival::finegray()`;  
+ Using `rcs::cph()` or `survival::coxph()` after expanding the data with `mstate::crprep()`;  
+ Using `riskRegression::FGR()` which is a formula interface for the function `crr` from the `cmprsk` package. 


#### Model development using survival package and finegray() function
```{r, finegray, fig.align='center',warning=FALSE}

rdata_fg <- finegray(Surv(time, status) ~ ., 
                     etype = "rec",
                     data = rdata)

fgfit <- coxph(Surv(fgstart, fgstop, fgstatus) ~ 
                 age + size + ncat + hr_status,
               weight = fgwt, 
               data = rdata_fg)


```

#### Model development using mstate package and crprep() function
```{r, mstate, fig.align='center',warning=FALSE}
# Expand data to prepare for fitting the model 
primary_event <- 1 # primary event is 1 (breast cancer recurrence)
# set 2 to fit a model for non-recurrence mortality.
rdata.w <- mstate::crprep(
  Tstop = "time",
  status = "status_num",
  trans = c(1, 2),
  id = "id",
  keep = c("age", "size", "ncat", "hr_status"),
  data = rdata
)
rdata.w1 <- rdata.w %>% filter(failcode == primary_event)

dd <- datadist(rdata)
options(datadist = "dd")
options(prType = "html")

fit_fg <- cph(Surv(Tstart, Tstop, status == 1) ~
                age + size + ncat + hr_status, 
                  weights = weight.cens,
                  x = T, 
                  y = T, 
                  surv = T, 
                  data = rdata.w1
)
print(fit_fg)
options(datadist = NULL)
```

#### Model development using riskRegression package and FGR() function

```{r, riskRegression, fig.align='center',warning=FALSE}
# We also fit the FG model using riskRegression::FGR()
# because it is useful to evaluate some of the prediction performance
# measures later

primary_event <- 1 # primary event is 1 (breast cancer recurrence)
# set 2 to fit a model for non-recurrence mortality.

fit_fgr <- FGR(Hist(time, status_num) ~ 
                 age + size +
                 ncat + hr_status, 
               cause = primary_event,
               data = rdata)
```

```{r, comp_coeff, fig.align='center',warning=FALSE}
res_coef <- cbind.data.frame( 
                             "survival package" = fgfit$coefficients,
                             "mstate + survival/rms packages" = 
                               fit_fg$coefficients,
                             "riskRegression package" = fit_fgr$crrFit$coef)
k <- 4
res_coef <- round(res_coef, k)
kable(res_coef) %>% 
  kable_styling("striped", position = "center")
```


The coefficients of the models indicated that higher size, positive nodal status and ER- and/or PR- were associated with higher risk to develop a breast cancer recurrence.


## Reproducibility ticket

```{r repro_ticket, echo=TRUE}
sessionInfo()
```
